---
title: "Homework3"
author: "Yunjia Liu"
date: "2024-10-14"
output: github_document
---

```{r setup, include=FALSE}
library(p8105.datasets)
library(tidyverse)
library(tidyr)
library(ggridges)
library(patchwork)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
```

## Problem1
Load the dataset.
```{r load data}
data("ny_noaa")
```

Answer to the questions.

1. Data Cleaning and Exploring the Snowfall Data

First, we separate variables for year, month, and day in the dataset. Besides, noticing the tmax and tmin are chr in the table, we will convert the temperature into the numeric. Also, according to the description of the dataset, temperature and precipitation are in tenths of degrees C and mm separately. So, we also need to convert the unit for these two variables. The unit of snowfall is already in mm.

```{r clean data}
ny_noaa_clean = ny_noaa |>
  separate(date, c('year', 'month', 'day'), sep = "-",remove = FALSE) |>
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         tmax = tmax / 10,
         tmin = tmin / 10,
         prcp = prcp / 10,
         month = as.numeric(month),
         year = as.numeric(year))
```

Second, explore the snowfall data. The most common snowfall value is 0 mm, which suggests many observations recorded no snowfall. This could also indicate periods of no snow or that snowfall isn’t always measured consistently. The second common value is NA, which suggests the measurement of snowfall is missing. The third and fourth common is 25 and 13 mm.

```{r}
# Snowfall summary
snowfall_counts = ny_noaa_clean |>
  count(snow) |>
  arrange(desc(n))

print(head(snowfall_counts, 10))

```

2. Two-Panel Plot: Average Max Temperature in January and July. We first calculate the average max temperature by grouping the date and filtering the months from 12 months to Jan and July. 

The plot shows the average maximum temperature in January and July across several weather stations in New York, with trends for each station shown across the years. These months represent two extreme seasons: Winter (cold) and Summer (hot). Temperatures in Jan is be much lower compared to July. All stations in NY seems to have similar temperature pattern in Jan/July in one specific year. There are some individual points(outliners) in the both panel that show unusual temperatures, which might indicate data entry errors or unusual weather patterns. 

```{r}
temp_summary = ny_noaa_clean |>
  group_by(id, year, month) |>
  filter(month %in% c(1, 7)) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE))


temp_summary |>
  ggplot(aes(x = year, y = mean_tmax, group = id, color = factor(month))) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  facet_wrap(~month, ncol = 1,labeller = labeller(month = c(`1` = "January", `7` = "July"))) +
  labs(
    title = "Average Max Temperature in January and July",
    x = "Year", y = "Average Max Temperature (°C)"
  ) +
  scale_x_continuous(
    breaks = seq(min(temp_summary$year), max(temp_summary$year), by = 3)
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

```

3. Two-Panel Plot: (i) Hex plot for tmax vs tmin and (ii) ridge plot for the distribution of snowfall

From the hex plot, we can find that there is a positive correlation between tmax and tmin since warmer days tend to have both higher minimum and maximum temperatures.

From the ridge plot, we will see that in most of all years many stations had snowfall between 0 and 35 mm. Additionally, it is obvious that certain stations see about 45 mm of snow and a small group of stations that sees nearly 80 mm. It is probably caused by miscalcualtion between unit systems. Also, we can find that there's lower snowfall each year from past to now, which may serve as a sign for a global warning.

```{r ,warning=FALSE}
tmax_tmin_plot = 
  ny_noaa_clean |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +
  labs(
    title = "tmax vs tmin",
    x = "Minimum Temperature (°C)", 
    y = "Maximum Temperature (°C)"
  ) +
  theme_minimal() +
    theme(legend.position = "bottom",
          legend.text = element_text(size=5),
          plot.title = element_text(hjust = 0.5))

snowfall_plot = 
  ny_noaa_clean |>
  filter(snow > 0 & snow < 100, !is.na(snow)) |>
  ggplot(aes(x = snow, y = factor(year))) +
  geom_density_ridges(scale = 1.2, rel_min_height = 0.01) +  
  scale_fill_viridis_c(option = "C") +  
  labs(
    title = "Distribution of Snowfall (0-100 mm) by Year",
    x = "Snowfall (mm)", y = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",  
    axis.text.y = element_text(size = 8)  
  )

print(tmax_tmin_plot + snowfall_plot)
```


## Problem2

1.1 Load the dataset.
Skip some rows in the beginning of nhanes_covar.csv to read the correct data.
```{r}
accel_data = read_csv("./data/nhanes_accel.csv") |>
  janitor::clean_names()  

demographic_data = read_csv("./data/nhanes_covar.csv",skip = 4) |>
  janitor::clean_names()
```

1.2 Cleaning the data.
We exclude participants under 21 and those with missing demographic data. Then, encoding sex as factor (originally double) and re-coding the education level.
```{r}
demographic_data_clean =
  demographic_data |>
  filter(age >= 21, !is.na(sex), !is.na(education)) |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    education =
      case_match(
        education,
        1 ~ "Less than high school",
        2 ~ "High school equivalent",
        3 ~ "More than high school"
      ),
    sex = as.factor(sex),
    education = as.factor(education)
  )
```

2. We merge Accelerometer Data and Demographic Data by participants id. And then create a table of men and female by education level. Finally, we visualize age distributions by sex and education level using density plot. The plot has age as x-axis and is divided into three categories by education level, colored using different colour for male and female.

```{r}
merged_data = demographic_data_clean |>
  inner_join(accel_data, by = "seqn")

education_table = merged_data |>
  group_by(education, sex) |>
  summarize(count = n(), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = count, values_fill = 0)
print(education_table)

merged_data |>
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~education, ncol = 2) +
  labs(
    title = "Age Distribution by Sex and Education Level",
    x = "Age", y = "Density"
  ) +
  theme_minimal() +
   theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))
```

3. Aggregate Total Activity for Each Participant and Plot against Age

Sum MIMS values for each participant to get total activity and then plot total activity against age, with panels for each education level and color by sex.
This plot shows the relationship between total activity (measured in MIMS) and age, with data split by sex and education level. The trend lines (using geom_smooth() with loess smoothing) indicate how activity changes with age, providing a way to compare activity levels between males and females within each education level.

Comment:

- Across all education levels, we observe a slight decline in total activity with *increasing age*.The loess curves show a subtle downward trend for both sexes in most panels, indicating that physical activity tends to decrease with age.
- Participants with more than *high school education* exhibit more stable activity patterns, suggesting that education may play a role in maintaining consistent physical activity over time.
- Participants with lower education levels show greater variability in their physical activity patterns, indicating that lifestyle factors may contribute to differences in physical activity within this group.
- While there are some small differences in specific education categories (e.g., slight male advantage at younger ages), the overall patterns for men and women are quite similar.

```{r}
total_activity = merged_data |>
  group_by(seqn, age, sex, education) |>
  summarize(total_activity = sum(c(min1,min1440), na.rm = TRUE), .groups = "drop")


total_activity |>
ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  facet_wrap(~education, ncol = 2) +
  labs(
    title = "Total Activity vs Age by Sex and Education Level",
    x = "Age", y = "Total Activity (MIMS)"
  ) +
  theme_minimal()+
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))
```

4. Create 24-hour monitor activity plot

We first pivot the dataset into a longer version so that we can link the exact time and mims for each individual.Also, we rename and mutate the exact time 'X' from the original column name 'minX', converting it into a numeric. After reshaping, each row now corresponds to a single participant’s activity for a given minute of the day. Then, we group and summarize the data to calculate the average activity (MIMS) for each minute, split by sex and education level.

Comment on the 24-hour monitor activity plot:

- Across all education levels, participants tend to be most active during the daytime and less active at night, reflecting normal sleep-wake cycles.
- Participants with higher education show slightly more consistent and higher peak activity levels, suggesting a possible link between education and more structured or healthier lifestyles.
- While small differences exist, overall the patterns are similar between men and women. Any differences are more noticeable among participants with less education, where men tend to show slightly higher activity during the day.

```{r}
merged_data_new = merged_data |>
  pivot_longer(min1:min1440,
               names_to='time',
               values_to='mims',
               names_prefix = 'min') |>
  rename(time_min=time)|>
  mutate(time_min=as.numeric(time_min))

hourly_activity = merged_data_new |>
  group_by(time_min, sex, education) |>
  summarize(mean_mims = mean(mims, na.rm = TRUE), .groups = "drop")

ggplot(hourly_activity, aes(x = time_min, y = mean_mims, color = sex)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  facet_wrap(~education, ncol = 1) +
  labs(
    title = "24-Hour Activity Patterns by Education Level and Sex",
    x = "Minute of Day", y = "Mean Activity (MIMS)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))  

```


## Problem 3

3.1 We first load the data first and combine the four datasets with year and month labeled. Then, we recode the varibale type of some columns.
```{r load adat in problem3}
jan_2020 = read_csv("./data/citibike/Jan 2020 Citi.csv") |>
  janitor::clean_names() |>
  mutate(year_month = "2020-01")

july_2020 = read_csv("./data/citibike/July 2020 Citi.csv")|>
  janitor::clean_names() |>
  mutate(year_month = "2020-07")

jan_2024 = read_csv("./data/citibike/Jan 2024 Citi.csv") |>
  janitor::clean_names() |>
  mutate(year_month = "2024-01")

july_2024 = read_csv("./data/citibike/July 2024 Citi.csv") |>
  janitor::clean_names() |>
  mutate(year_month = "2024-07")

citibike_data = 
  bind_rows(jan_2020, july_2020, jan_2024, july_2024) |>
  separate(year_month, into = c('year','month'), sep = '-') |>
  mutate(
    year = as.numeric(year),
    month = as.numeric(month),
    month = 
      case_match(
        month, 
        1 ~ "Jan", 
        7 ~ "July"),
    rideable_type = factor(rideable_type),
    member_casual = factor(member_casual)
  ) 
```


3.2 Create a table of total rides by year, month, and membership type.

The table provides counts of Citi Bike rides grouped by year, month, and membership type (casual vs. member). Here is the comment on the table:

- Rides peak in July (summer) and drop in January (winter).
- Members consistently take more rides than casual users.
- Both member and casual usage increased significantly between 2020 and 2024.
- Casual usage grows more in July, reflecting increased tourism and outdoor activities in summer.

This analysis highlights the importance of seasonality and membership programs in Citi Bike's operations and suggests that service improvements have driven growth in ridership over the years.
```{r}
rides_table = 
  citibike_data |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n(), .groups = "drop") |>
  pivot_wider(names_from = member_casual, values_from = total_rides, values_fill = 0)

print(rides_table)
```


3.3 Filter data for July 2024 and count rides by starting station.

This table shows the five most popular starting stations in the Citi Bike system for July 2024, along with the total number of rides originating from each station.

```{r}
top_stations = citibike_data |>
  filter(year == 2024, month == "July") |>
  group_by(start_station_name) |>
  summarize(total_rides = n(), .groups = "drop") |>
  arrange(desc(total_rides)) |>
  slice_head(n = 5)

print(top_stations)

```

3.4 Plot the Effect of Day, Month, and Year on Median Ride Duration.

We calculate median ride duration by day of week, month, and year. This plot shows how median ride duration (in minutes) varies by day of the week, with separate panels for January and July. The color scale at the bottom reflects different years (from 2020 to 2024). Here's the comment on the plot:


```{r}
ride_duration_stats = citibike_data |>
  group_by(year, month, weekdays) |>
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = "drop")

ggplot(ride_duration_stats, aes(x = weekdays, y = median_duration, color = year)) +
  geom_point() +
  geom_line(aes(group = interaction(year, month))) +
  facet_wrap(~month, ncol = 2) +
  labs(
    title = "Effect of Day, Month, and Year on Median Ride Duration",
    x = "Day of the Week", y = "Median Ride Duration (Minutes)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

3.5 Investigate the Impact of Month, Membership, and Bike Type on Ride Duration in 2024.
Filter data for 2024 and create a plot of ride duration distributions

```{r}
# Filter data for 2024 and create a plot of ride duration distributions
ggplot(citibike_data %>% filter(year == 2024), 
       aes(x = duration, fill = member_casual)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.6) +
  facet_grid(rideable_type ~ month) +
  labs(
    title = "Distribution of Ride Duration by Month, Membership, and Bike Type (2024)",
    x = "Ride Duration (Minutes)", y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))

```

