---
title: "Homework3"
author: "Yunjia Liu"
date: "2024-10-14"
output: github_document
---

```{r setup, include=FALSE}
library(p8105.datasets)
library(tidyverse)
library(tidyr)
library(ggridges)
```

## Problem1
Load the dataset.
```{r load data}
data("ny_noaa")
```

Answer to the questions.
First, we separate variables for year, month, and day in the dataset. Besides, noticing the tmax and tmin are chr in the table, we will convert the temperature into the numeric. Also, according to the description of the dataset, temperature and precipitation are in tenths of degrees C and mm separately. So, we also need to convert the unit for these two variables. The unit of snowfall is already in mm.

```{r clean data}
ny_noaa_clean = ny_noaa |>
  separate(date, c('year', 'month', 'day'), sep = "-",remove = FALSE) |>
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         tmax = tmax / 10,
         tmin = tmin / 10,
         prcp = prcp / 10,
         month = as.numeric(month),
         year = as.numeric(year))
```

Second, explore the snowfall data. The most common snowfall value is 0 mm, which suggests many observations recorded no snowfall. This could also indicate periods of no snow or that snowfall isn’t always measured consistently. The second common value is NA, which suggests the measurement of snowfall is missing. The third and fourth common is 25 and 13 mm.

```{r}
# Snowfall summary
snowfall_counts = ny_noaa_clean |>
  count(snow) |>
  arrange(desc(n))

print(head(snowfall_counts, 10))

```

Two-Panel Plot: Average Max Temperature in January and July. We first calculate the average max temperature by grouping the date and filtering the months from 12 months to Jan and July. 

The plot shows the average maximum temperature in January and July across several weather stations in New York, with trends for each station shown across the years. These months represent two extreme seasons: Winter (cold) and Summer (hot). Temperatures in Jan is be much lower compared to July. There are some individual points(outliners) in the both panel that show unusual temperatures, which might indicate data entry errors or unusual weather patterns.

```{r}
temp_summary = ny_noaa_clean |>
  group_by(id, year, month) |>
  filter(month %in% c(1, 7)) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE))


temp_summary |>
  ggplot(aes(x = year, y = mean_tmax, group = id, color = factor(month))) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  facet_wrap(~month, ncol = 1,labeller = labeller(month = c(`1` = "January", `7` = "July"))) +
  labs(
    title = "Average Max Temperature in January and July",
    x = "Year", y = "Average Max Temperature (°C)"
  ) +
  scale_x_continuous(
    breaks = seq(min(temp_summary$year), max(temp_summary$year), by = 3)
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

```




