---
title: "Homework3"
author: "Yunjia Liu"
date: "2024-10-14"
output: github_document
---

```{r setup, include=FALSE}
library(p8105.datasets)
library(tidyverse)
library(tidyr)
library(ggridges)
library(patchwork)
```

## Problem1
Load the dataset.
```{r load data}
data("ny_noaa")
```

Answer to the questions.
1. Data Cleaning and Exploring the Snowfall Data
First, we separate variables for year, month, and day in the dataset. Besides, noticing the tmax and tmin are chr in the table, we will convert the temperature into the numeric. Also, according to the description of the dataset, temperature and precipitation are in tenths of degrees C and mm separately. So, we also need to convert the unit for these two variables. The unit of snowfall is already in mm.

```{r clean data}
ny_noaa_clean = ny_noaa |>
  separate(date, c('year', 'month', 'day'), sep = "-",remove = FALSE) |>
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         tmax = tmax / 10,
         tmin = tmin / 10,
         prcp = prcp / 10,
         month = as.numeric(month),
         year = as.numeric(year))
```

Second, explore the snowfall data. The most common snowfall value is 0 mm, which suggests many observations recorded no snowfall. This could also indicate periods of no snow or that snowfall isn’t always measured consistently. The second common value is NA, which suggests the measurement of snowfall is missing. The third and fourth common is 25 and 13 mm.

```{r}
# Snowfall summary
snowfall_counts = ny_noaa_clean |>
  count(snow) |>
  arrange(desc(n))

print(head(snowfall_counts, 10))

```

2. Two-Panel Plot: Average Max Temperature in January and July. We first calculate the average max temperature by grouping the date and filtering the months from 12 months to Jan and July. 

The plot shows the average maximum temperature in January and July across several weather stations in New York, with trends for each station shown across the years. These months represent two extreme seasons: Winter (cold) and Summer (hot). Temperatures in Jan is be much lower compared to July. All stations in NY seems to have similar temperature pattern in Jan/July in one specific year. There are some individual points(outliners) in the both panel that show unusual temperatures, which might indicate data entry errors or unusual weather patterns. 

```{r}
temp_summary = ny_noaa_clean |>
  group_by(id, year, month) |>
  filter(month %in% c(1, 7)) |>
  summarize(mean_tmax = mean(tmax, na.rm = TRUE))


temp_summary |>
  ggplot(aes(x = year, y = mean_tmax, group = id, color = factor(month))) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  facet_wrap(~month, ncol = 1,labeller = labeller(month = c(`1` = "January", `7` = "July"))) +
  labs(
    title = "Average Max Temperature in January and July",
    x = "Year", y = "Average Max Temperature (°C)"
  ) +
  scale_x_continuous(
    breaks = seq(min(temp_summary$year), max(temp_summary$year), by = 3)
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

```

3. Two-Panel Plot: (i) Hex plot for tmax vs tmin and (ii) ridge plot for the distribution of snowfall

From the hexplot, we can find that there is a positive correlation between tmax and tmin since warmer days tend to have both higher minimum and maximum temperatures.

From the ridge plot, we will see that in most of all years
```{r ,warning=FALSE}
tmax_tmin_plot = 
  ny_noaa_clean |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +
  labs(
    title = "tmax vs tmin",
    x = "Minimum Temperature (°C)", 
    y = "Maximum Temperature (°C)"
  ) +
  theme_minimal() +
    theme(legend.position = "bottom",
          legend.text = element_text(size=5),
          plot.title = element_text(hjust = 0.5))

snowfall_plot = 
  ny_noaa_clean |>
  filter(snow > 0 & snow < 100, !is.na(snow)) |>
  ggplot(aes(x = snow, y = factor(year))) +
  geom_density_ridges(scale = 1.2, rel_min_height = 0.01) +  
  scale_fill_viridis_c(option = "C") +  
  labs(
    title = "Distribution of Snowfall (0-100 mm) by Year",
    x = "Snowfall (mm)", y = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",  
    axis.text.y = element_text(size = 8)  
  )

print(tmax_tmin_plot + snowfall_plot)
```


## Problem2

Load the dataset.
Skip some rows in the beginning of nhanes_covar.csv to read the correct data.
```{r}
accel_data = read_csv("./data/nhanes_accel.csv") |>
  janitor::clean_names()  

demographic_data = read_csv("./data/nhanes_covar.csv",skip = 4) |>
  janitor::clean_names()
```

Cleaning the data.
We exclude participants under 21 and those with missing demographic data. Then, encoding sex as factor (originally double) and re-coding the education level.
```{r}
demographic_data_clean =
  demographic_data |>
  filter(age >= 21, !is.na(sex), !is.na(education)) |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    education =
      case_match(
        education,
        1 ~ "Less than high school",
        2 ~ "High school equivalent",
        3 ~ "More than high school"
      ),
    sex = as.factor(sex),
    education = as.factor(education)
  )
```

We merge Accelerometer Data and Demographic Data by participants id. And then create a table of men and female by education level. Finally, we visualize age distributions by sex and education level.

```{r}
merged_data = demographic_data_clean |>
  inner_join(accel_data, by = "seqn")

education_table = merged_data |>
  group_by(education, sex) |>
  summarize(count = n(), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = count, values_fill = 0)
print(education_table)

merged_data |>
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~education, ncol = 2) +
  labs(
    title = "Age Distribution by Sex and Education Level",
    x = "Age", y = "Density"
  ) +
  theme_minimal() +
   theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5))

```
